---
title: "CIRI3 decoupled-vs-joint validation report"
author: "CIRI3 regression harness"
date: "`r format(Sys.time(), '%Y-%m-%d %H:%M')`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
    code_folding: hide
params:
  out_root: "/pastel/tools/circRNA_tools/test_data/decoupled_comparison"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 8, fig.height = 6)
library(tidyverse)
library(knitr)
```

This report reads the outputs produced by
`scripts/test_decoupled_pipeline_star_fullsize.sh` (or the BWA variant)
and compares the joint `-W 1` pipeline's BSJ/FSJ matrices against the
decoupled `SCAN1 → BUILD_UNIVERSE → SCAN2 → FINALIZE` matrices.

To rerender after a new test run:

```
Rscript -e "rmarkdown::render('scripts/validation_report.Rmd')"
```

or, to point at a different output directory:

```
Rscript -e "rmarkdown::render('scripts/validation_report.Rmd', params=list(out_root='/path/to/out_root'))"
```

---

## Test configuration

```{r paths}
out_root        <- params$out_root
orig_dir        <- file.path(out_root, "original")
decoupled_dir   <- file.path(out_root, "decoupled")
finalize_dir    <- file.path(decoupled_dir, "finalize")
bench_dir       <- file.path(out_root, "bench")

required <- c(
  "original/result.BSJ_Matrix",
  "original/result.FSJ_Matrix",
  "decoupled/finalize/result.BSJ_Matrix",
  "decoupled/finalize/result.FSJ_Matrix"
)
missing <- required[!file.exists(file.path(out_root, required))]
if (length(missing) > 0) {
  stop("Missing expected input files:\n  ",
       paste(file.path(out_root, missing), collapse = "\n  "))
}
tibble(key = c("out_root", "orig_dir", "decoupled_dir"),
       value = c(out_root, orig_dir, decoupled_dir)) |>
  kable(caption = "Paths used by this report")
```

## Load matrices

```{r load}
joint_BSJ     <- read_tsv(file.path(orig_dir, "result.BSJ_Matrix"))
joint_FSJ     <- read_tsv(file.path(orig_dir, "result.FSJ_Matrix"))
decoupled_BSJ <- read_tsv(file.path(finalize_dir, "result.BSJ_Matrix"))
decoupled_FSJ <- read_tsv(file.path(finalize_dir, "result.FSJ_Matrix"))

# Joint writes sample columns by input-file basename (e.g. "bwa.sam");
# decoupled uses the sampleName column from finalize_samples.tsv. They mean
# the same thing in the same order, so align positionally using the decoupled
# names — those are more informative.
align_cols <- function(df, ref_names) {
  stopifnot(ncol(df) == length(ref_names) + 1)
  colnames(df) <- c("circRNA_ID", ref_names)
  df
}
ref_names    <- colnames(decoupled_BSJ)[-1]
joint_BSJ    <- align_cols(joint_BSJ,    ref_names)
joint_FSJ    <- align_cols(joint_FSJ,    ref_names)

tibble(
  matrix     = c("BSJ joint", "BSJ decoupled", "FSJ joint", "FSJ decoupled"),
  nrow       = c(nrow(joint_BSJ), nrow(decoupled_BSJ),
                 nrow(joint_FSJ), nrow(decoupled_FSJ)),
  n_samples  = length(ref_names)
) |> kable(caption = "Matrix dimensions (sample columns aligned positionally)")

cat("Sample column names used:\n")
cat(paste0("  ", ref_names, collapse = "\n"), "\n")
```

## circRNA set agreement

```{r sets}
both   <- intersect(joint_BSJ$circRNA_ID, decoupled_BSJ$circRNA_ID)
only_j <- setdiff(joint_BSJ$circRNA_ID, decoupled_BSJ$circRNA_ID)
only_d <- setdiff(decoupled_BSJ$circRNA_ID, joint_BSJ$circRNA_ID)

tibble(
  set = c("circRNAs in both pipelines",
          "joint only",
          "decoupled only",
          "Jaccard (intersection / union)"),
  n   = c(length(both),
          length(only_j),
          length(only_d),
          round(length(both) /
                  length(union(joint_BSJ$circRNA_ID,
                               decoupled_BSJ$circRNA_ID)), 4))
) |> kable(caption = "circRNA ID set agreement")
```

### Sample joint-only IDs (first 20)

```{r}
head(only_j, 20) |> as_tibble() |> kable(col.names = "joint-only circRNA_ID")
```

### Sample decoupled-only IDs (first 20)

```{r}
head(only_d, 20) |> as_tibble() |> kable(col.names = "decoupled-only circRNA_ID")
```

## BSJ matrix agreement

```{r bsj_long}
bsj_merged <- full_join(joint_BSJ, decoupled_BSJ,
                        by = "circRNA_ID",
                        suffix = c(".joint", ".decoupled"))

bsj_long <- bsj_merged |>
  pivot_longer(cols = -circRNA_ID,
               names_to = c("Sample", "Run"),
               names_pattern = "(.*)\\.(joint|decoupled)$") |>
  pivot_wider(names_from = Run, values_from = value) |>
  replace_na(list(joint = 0, decoupled = 0))

bsj_long |>
  group_by(Sample) |>
  summarise(
    n_rows            = n(),
    exactly_equal     = sum(joint == decoupled),
    joint_greater     = sum(joint > decoupled),
    decoupled_greater = sum(decoupled > joint),
    max_abs_diff      = max(abs(joint - decoupled)),
    pct_exact         = round(100 * mean(joint == decoupled), 2)
  ) |> kable(caption = "BSJ matrix: per-sample cell-level agreement")
```

### BSJ scatter: joint vs. decoupled

```{r bsj_scatter}
ggplot(bsj_long, aes(x = joint, y = decoupled)) +
  geom_point(alpha = 0.4, size = 0.6) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  facet_wrap(~ Sample, scales = "free") +
  labs(x = "Joint pipeline BSJ count",
       y = "Decoupled pipeline BSJ count",
       title = "BSJ counts per circRNA per sample") +
  theme_bw()
```

### BSJ cell-level discordance (top 20)

```{r bsj_top}
bsj_long |>
  mutate(abs_diff = abs(joint - decoupled)) |>
  filter(abs_diff > 0) |>
  arrange(desc(abs_diff)) |>
  head(20) |> kable(caption = "Largest per-cell BSJ disagreements")
```

## FSJ matrix agreement

```{r fsj_long}
fsj_merged <- full_join(joint_FSJ, decoupled_FSJ,
                        by = "circRNA_ID",
                        suffix = c(".joint", ".decoupled"))
fsj_long <- fsj_merged |>
  pivot_longer(cols = -circRNA_ID,
               names_to = c("Sample", "Run"),
               names_pattern = "(.*)\\.(joint|decoupled)$") |>
  pivot_wider(names_from = Run, values_from = value) |>
  replace_na(list(joint = 0, decoupled = 0))

fsj_long |>
  group_by(Sample) |>
  summarise(
    n_rows            = n(),
    exactly_equal     = sum(joint == decoupled),
    joint_greater     = sum(joint > decoupled),
    decoupled_greater = sum(decoupled > joint),
    max_abs_diff      = max(abs(joint - decoupled)),
    pct_exact         = round(100 * mean(joint == decoupled), 2)
  ) |> kable(caption = "FSJ matrix: per-sample cell-level agreement")
```

### FSJ scatter: joint vs. decoupled

```{r fsj_scatter}
ggplot(fsj_long, aes(x = joint, y = decoupled)) +
  geom_point(alpha = 0.4, size = 0.6) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  facet_wrap(~ Sample, scales = "free") +
  labs(x = "Joint pipeline FSJ count",
       y = "Decoupled pipeline FSJ count",
       title = "FSJ counts per circRNA per sample") +
  theme_bw()
```

### FSJ cell-level discordance (top 20)

```{r fsj_top}
fsj_long |>
  mutate(abs_diff = abs(joint - decoupled)) |>
  filter(abs_diff > 0) |>
  arrange(desc(abs_diff)) |>
  head(20) |> kable(caption = "Largest per-cell FSJ disagreements")
```

## Detection-correlation per sample

Per-sample Pearson and Spearman correlation between joint and decoupled
counts — a value of 1.00 means per-cell agreement is perfect; values
under ~0.99 flag a real divergence.

```{r corr}
sample_corr <- function(df, what) {
  df |>
    group_by(Sample) |>
    summarise(
      pearson  = round(cor(joint, decoupled, method = "pearson"),  4),
      spearman = round(cor(joint, decoupled, method = "spearman"), 4)
    ) |>
    mutate(matrix = what, .before = 1)
}
bind_rows(sample_corr(bsj_long, "BSJ"),
          sample_corr(fsj_long, "FSJ")) |>
  kable(caption = "Joint vs. decoupled correlation (per sample)")
```

## Benchmark (if present)

```{r bench}
bench_files <- list.files(bench_dir, pattern = "\\.bench$", full.names = TRUE)
if (length(bench_files) == 0) {
  cat("No benchmark files found in", bench_dir, "\n")
} else {
  parse_bench <- function(f) {
    txt <- readLines(f, warn = FALSE)
    if (length(txt) == 0) return(NULL)
    kv <- strsplit(txt[1], " ")[[1]]
    as_tibble(setNames(
      lapply(kv, function(x) strsplit(x, "=")[[1]][2]),
      sapply(kv, function(x) strsplit(x, "=")[[1]][1])
    )) |> mutate(stage = sub("\\.bench$", "", basename(f)), .before = 1)
  }
  bench_df <- bind_rows(lapply(bench_files, parse_bench))
  # Best-effort numeric coercion
  for (col in c("wall_sec", "user_sec", "sys_sec", "max_rss_kb")) {
    if (col %in% names(bench_df)) {
      bench_df[[col]] <- suppressWarnings(as.numeric(bench_df[[col]]))
    }
  }
  bench_df <- bench_df |> arrange(stage)
  if ("max_rss_kb" %in% names(bench_df)) {
    bench_df$max_rss_gb <- round(bench_df$max_rss_kb / 1024 / 1024, 2)
  }
  bench_df |> kable(caption = "Per-stage benchmark")
}
```

---

## Session info

```{r sessioninfo}
sessionInfo()
```
